A Systematic Literature Review on Extensions to Role Based Access Control

### 2013-02-27 paper/sections/abstract.tex
Updates to paper list into appendix. Adds structured abstract. Moves paper counts into process.

Since the introduction of the Role-based access control (RBAC) standard in the late 1990's, 
RBAC has become an increasingly popular access control mechanism for enterprises such as 
web applications, mobile computing and databases. RBAC restricts access to resources 
based on the identity of subjects connected to logical groupings of permissions called roles.  
The introduction of computing into new domains has led to proposals for extensions to the standard
RBAC model; for example, defining additional constraints among roles, adding context 
or defining new hierarchy relationships.  \textit{The goal of this work is to provide practioners an assessment of the state of extended models of RBAC and researchers with insights into the lack of robust evaluation of RBAC extension models.}
We conducted a systematic literature review by collecting and synthesizing relevant research
papers in the area of RBAC extensions. We initially collect XXXX papers from sources such as IEEE and ACM 
websites and selected X primary sources.  We performed a comparative analysis of the primary sources to 
find relationships among extended models and analyze the state of RBAC extension evaluations.

>>>>

\noindent
{\bf Context}: Since the introduction of the role based access control (RBAC) standard in the late 1990's, 
RBAC has become an increasingly popular access control mechanism for applications such as 
web applications, mobile computing and databases. RBAC restricts access to resources 
based on the identity of subjects connected to logical groupings of permissions called roles.  
The introduction of computing into new domains has led to proposals for extensions to the standard
RBAC model; for example, defining additional constraints among roles, adding context or defining new heirarchy relationships.
\\
\\
\noindent
{\bf Aim}: \textit{The goal of this work is to provide practioners an assessment of the state of extended models of RBAC and researchers with insights into the lack of robust evaluation of RBAC extension models.}
\\
\\
\noindent
{\bf Method}: We conducted a systematic literature review by collecting and synthesizing relevant research
papers in the area of RBAC extensions. We initially collect XXXX papers from sources such as IEEE and ACM 
websites and selected X primary sources.  We performed a comparative analysis of the primary sources to 
find relationships among extended models and analyze the state of RBAC extension evaluations.
\\
\\
\noindent
{\bf Results}:
\\
\\
\noindent
{\bf Conclusions}: We conducted a systematic literature review by collecting and synthesizing relevant research
papers in the area of RBAC extensions. We initially collect XXXX papers from sources such as IEEE and ACM 
websites and selected X primary sources.  We performed a comparative analysis of the primary sources to 
find relationships among extended models and analyze the state of RBAC extension evaluations.

### 2013-02-27 paper/sections/analysis.tex
Updates to paper list into appendix. Adds structured abstract. Moves paper counts into process.

What are the deficiencies in the current RBAC model? \\

Core RBAC model provides an abstraction of authorization model based on user role assignment (UA) and the role
permission assignment (PA). Since this simple model may not provide a fine-grained access control for sophisticated
security mechanism, a variety of extended RBAC models have been proposed over the years to meet
security requirements. Our surveyed 29 papers address more than one deficiencies described as follows.

\begin{itemize}
%	\item Core RBAC model needs additional contextual information and constraints to develop fine-grained policies in practice.
	\item Core RBAC does not incorporate context. Therefore, RBAC belongs to static access control model, which may not capture changes in environments.
	\item Core RBAC does not support for various constraints such as temporal and spatial constraints to design sophisticated policies on demand.
	\item Core RBAC does not provide an abstraction to additional user-defined attributes	(e.g., task and team) and their association with existing attributes.
	\item Core RBAC has limitation on delegation and role hierarchy. For example, partial inheritance in role hierarchy needs to be developed.  
\end{itemize}

@@

When administrators design a model, it is important to capture an important abstraction to help the model to be enforced in a system.
Administrators improve existing core RBAC model due to following reasons:

\begin{itemize}
	\item Model needs to incorporate additional contextual information and constraints to develop fine-grained policies in practice.
	\item Model needs to incorporate additional attributes such as
	task, team, purpose, organizational roles and collaborative activities over existing attributes. Moreover, new association between attributes
	are introduced. 
		\item Model needs to improve existing features such delegation and role hierarchy to meet fine-grained access control. 
\end{itemize}

@@

Do the extension models have corresponding implementations in practice? \\

@@

How are extended RBAC models evaluated in theory and in practice? \\

@@

\begin{itemize}
\item Time-based Performance
\item Complexity analysis
\item Comparison to standard RBAC
\item Mathematical accuracy
\item Example scenarios of the model in action
\item Experimental analysis of the model
\item Case study of the model in practice
\end{itemize}

@@

@@

What domains or scenarios serve as inspiration for these extensions? \\

>>>>

What categorizations exist within RBAC extensions? \\

\begin{itemize}

  \item Constraint:

  \item Context:

  \item Organizational:

  \item Privacy:

  \item Resource:

  \item Spatial:  Spatial constraints are location-based constraints in specifying access
	control policies. For example, in organizations, 	locations are enforced while a
	specific role is permitted to conduct an action. Consider that part-time employee works only in specific location.
	In such cases, the part-time employee role should access required resources only when the user is in the location. 
	Spatial constraints can incorporate either on roles, user-role assignments, or role-permission assignments.

  \item Spatio-Temporal:

  \item Task: A task is a fundamental unit of a business activity. Different from core RBAC, in task-role-based access control model, roles are not directly associated with permissions. Roles are first associated with tasks, which are associated permissions. For example, the employee role is associated with a task, which is to write a report. Then, this task is associated with a permission.

  \item Temporal:  Temporal constraints are time-based constraints in specifying access
	control policies. For example, in organizations, 	periodic temporal durations are enforced while a
	specific role is permitted to conduct an action. Consider that part-time employee works only from 9:00 a.m. to 3:00 p.m.
	In such cases, the part-time employee role should access required resources during the interval. 
	Temporal constraints can incorporate either on roles, user-role assignments, or role-permission assignments.

\end{itemize}

@@

Core RBAC model provides an abstraction of authorization model based on user role assignment (UA) and the role
permission assignment (PA). Since this simple model may not provide a fine-grained access control for sophisticated
security mechanism, a variety of extended RBAC models have been proposed over the years to meet
security requirements. When administrators design a model, it is important to capture an important abstraction to help the model to be enforced in a system.

\begin{itemize}
\item Core RBAC model needs additional contextual information and constraints to develop fine-grained policies in practice.
\item Core RBAC does not incorporate context. Therefore, RBAC belongs to static access control model, which may not capture changes in environments.
\item Core RBAC does not support for various constraints such as temporal and spatial constraints to design sophisticated policies on demand.
\item Core RBAC does not provide an abstraction to additional user-defined attributes	(e.g., task and team) and their association with existing attributes.
\item Core RBAC has limitation on delegation and role hierarchy. For example, partial inheritance in role hierarchy needs to be developed.  
\item Model needs to incorporate additional contextual information and constraints to develop fine-grained policies in practice.
\item Model needs to incorporate additional attributes such as
task, team, purpose, organizational roles and collaborative activities over existing attributes. Moreover, new association between attributes
are introduced. 
\item Model needs to improve existing features such delegation and role hierarchy to meet fine-grained access control. 
\end{itemize}

@@

Do the extension models have corresponding implementations? \\

@@

How are extension to RBAC evaluated theoritically and in practice? \\

@@

\begin{itemize}
\item Time-based Performance \cite{ni2010privacy}, \cite{aich09:role}
\item Complexity analysis \cite{bao08:role}, \cite{zhang06:collaborative}, \cite{chen08:spatio-temporal}, \cite{aich09:role}
\item Comparison to standard RBAC \cite{bao08:role}, \cite{zou2009crbac}, \cite{zhang06:collaborative}, \cite{zhao2008flexible}, \cite{ray07:spatio}
\item Mathematical modeling \cite{damian2007geo}, \cite{hansen2003spatial}, \cite{aich07:STARBAC}, \cite{chen08:spatio-temporal}, \cite{joshi05:generalized}
\item Example scenarios of the model in action \cite{alam06:constraint}, \cite{tzelepi01:flexible}, \cite{cholewka00:acontext-sensitive}, \cite{huang06:pervasive}, \cite{bao08:role}, \cite{jian2008extended}, \cite{yamazaki104:designing}, \cite{zou2009crbac}, \cite{ray07:spatio}, \cite{samuel07:spatio-temporal}, \cite{ray07:spatio}, \cite{joshi05:generalized}, \cite{yao2008task}, \cite{zhou2007team}, \cite{oh2003task}
\item Experimental analysis of the model
\item Case study of the model in practice \cite{motta03:contextual}
\end{itemize}

@@

TODO: Add references to list above.

@@

What domains have RBAC extensions been created for? \\

### 2013-02-27 paper/sections/process.tex
Updates to paper list into appendix. Adds structured abstract. Moves paper counts into process.

The systematic literature review process was developed in whole prior to application and agreed upon by the researchers following recommendations from Kitchenham's suggested processes \cite{kitchenham2007guidelines}.  The systematic literature review process was broken down in to four stages and the rest of this section is broken down by each stage.  The stages were:

\begin{itemize}
\item Development of a search strategy
\item Elimination of papers based on title
\item Elimination of papers based on abstract
\item Elimination of papers based on content and matching to elimination criteria
\end{itemize}

\subsection{Search Strategy}

For the first phase of our systematic literature review, a search strategy for finding papers was developed.  The search strategy was executed by an automated comprehensive search taking as input a set of academic search engines and a list of search criteria.  The search performed was done in an automated way using a set of scripts to query and collect data from each search engine with the criteria string as input.  For each criteria for each search engine, the results were captured until a stopping criteria was met.  The search algorithm was performed as follows:

\begin{enumerate}
\item Remote call to search engine with current search start position and the current search criteria.
\item Parse results and extract paper title, authors and year of publication.
\item Compare results against stopping criteria.
	\begin{itemize}
	\item If stopping criteria met, stop search.
	\item If stopping criteria not met, increase search position by number of results and return to step 1.
	\end{itemize}
\end{enumerate}

The stopping criteria used was either after the first 1000 results, a limitation imposed by some of the search engines, or if ten consecutive results did not contain the search criteria phrase within the title.  After gathering all 12 data sets, the data was combined into a master list by systematically comparing the bibliographic information for each.  After producing a master list, a series of elimination rounds were performed to narrow the list of papers and identify primary sources.

\subsection{Elimination Rounds}

@@

\begin{itemize}
\item Each reviewer independently classified papers as relevant, irrelevant or uncertain based off elimination criteria
\item Those papers marked as relevant by both reviewers were kept and those marked irrelevant by both were thrown out.
\item Papers marked as relevant, or irrelevant by a single reviewer were combined with all papers marked as uncertain and discussed by both reviewers.  From this discussion, papers were either thrown out or kept until the next round of the review.  Ties were broken by an indepedent party.
\end{itemize}

\subsubsection{Title Elimination}

@@

\subsubsection{Abstract Elimination}

@@

\subsubsection{Content Elimination}

@@

\begin{itemize}
\item Does this model extend the core model?
\item What reasons and evidence do researchers give that RBAC needs extension?
\item Was the paper and subsequent model inspired by a real world example?
\item Is there any evaluation of the proposed model? If yes, how did they do one? If no, why?
\item Did the authors implement their model?
\end{itemize}

>>>>

The authors adopted and applied a systematic literature review following recommendations from Kitchenham's suggested processes \cite{kitchenham2007guidelines}.  The systematic literature review process was broken down in to four stages and the rest of this section is broken down by each stage.  The stages were:

\begin{itemize}
\item Step 1: Development of a search strategy
\item Step 2: Elimination of papers based on title criteria
\item Step 3: Elimination of papers based on abstract criteria
\item Step 4: Elimination of papers based on content and matching to elimination criteria
\end{itemize}

\subsection{Step 1: Search Strategy}

For the first phase of our systematic literature review, the authors developed a search strategy for finding papers.  The search strategy was executed by an automated comprehensive search taking as input a set of academic search engines and a list of search terms.  The search was performed by applying each search term to each engine incrementally until a stopping criteria was met. Table \ref{tab:search_results} lists the four search engines along the left most column with the three search term across the top row along with the papers for each criteria and engine combination.  The search algorithm was performed as follows:

\begin{enumerate}
\item Call to search engine with current search position and current search term.
\item Parse results and extract paper title, authors and year of publication.
\item Compare results against stopping criteria:
	\begin{itemize}
	\item If the size of the result set is greater than or equal to 1000 then stop.
	\item If the last ten results did not contain the search term phrase within the tite then stop.
	\end{itemize}
\item If stopping criteria not met, increment search position and go back to step one.
\end{enumerate}

The result set size stopping criteria was chosen due to a technical limitation of some search engines.  The stopping criteria related to the last ten titles is meant to stop after relevant results are no longer being returned by the search engine.  After gathering all 12 data sets, the authors combined the papers into a master list by systematically comparing the bibliographic information for each.  The master list of 1716 papers was used to perform a series of elimination rounds to narrow the list of papers and identify primary sources.

\begin{table}
\centering
\begin{tabular}{|p{3.5cm}|p{1.25cm}|p{4.25cm}|p{4.25cm}|p{1cm}|}

\hline
 & 
\textbf{RBAC} & 
\textbf{role based access control} & 
\textbf{role-based access control} & 
\textbf{Total}
\\\hline

Google Scholar & 651 & 213 & 435 & 1299 \\\hline
ACM Portal & 500 & 20 & 720 & 1240 \\\hline
IEEExplore & 200 & 40 & 230 & 470 \\\hline
CiteSeerX & 100 & 100 & 150 & 350 \\\hline
 &  &  &  & \\\hline
Totals & 1451 & 373 & 1535 & 3359 \\\hline
Combined &  &  &  & \textbf{1716} \\\hline

\end{tabular}
\caption{Paper counts after applying search strategy}
\label{tab:search_results}
\end{table}

TODO: Add elimination stats.

\subsection{Steps 2-4: Elimination Rounds}

@@

\begin{itemize}
\item The two first authors independently classified papers as relevant, irrelevant or uncertain based off elimination criteria
\item Those papers marked as relevant by both reviewers were kept and those marked irrelevant by both were thrown out.
\item Papers marked as relevant, or irrelevant by a single reviewer were combined with all papers marked as uncertain and discussed by both reviewers.  From this discussion, papers were either thrown out or kept until the next round of the review.
\end{itemize}

\subsubsection{Step 2: Title Elimination}

@@

\subsubsection{Step 3: Abstract Elimination}

@@

\subsubsection{Step 4: Content Elimination}

@@

\begin{enumerate}
\item Does this model extend the core model? (Exclusion)
\item Do the researchers give evidence that RBAC needs extension? (Inclusion)
\item Was the paper and subsequent model inspired by a real world example?  (Conditional Inclusion)
\item Did the researchers offer any evaluation of the proposed model? If yes, how did they do one? If no, why? (Conditional Inclusion)
\item Did the authors implement their model? (Inclusion)
\end{enumerate}

Question 1 was a definitive exclusion criteria as any paper that failed in the affirmative was rejected.  Questions 3 and 4 were marked as conditional includes given that they were connected in making a decision.  A paper that met Question 3 but not 4, or met 4 but not 3 would be included since in some cases the real world examples served as research evaluations and without this conditional include the paper list size would be too small to be significant.

### 2013-02-27 paper/sections/results.tex
Updates to paper list into appendix. Adds structured abstract. Moves paper counts into process.

For our systematic literature review, the search strategy was applied to the following search engines based on popularity and database size:

\begin{itemize}
\item Google Scholar - \url{http://scholar.google.com}
\item IEEExplore - \url{http://ieeexplore.ieee.org}
\item ACM Portal - \url{http://dl.acm.org}
\item CiteSeerX - \url{http://http://citeseerx.ist.psu.edu/index}
\end{itemize}

For each of the criteria below, a search was performed using the exact query for each of the search engines for a total of 12 data sets.  The criteria used were:

\begin{itemize}
\item role based access control
\item RBAC
\item role-based access control
\end{itemize}

The results of the search are summarised below:

\begin{center}
\begin{tabular}{|p{3.5cm}|p{1.25cm}|p{4.25cm}|p{4.25cm}|p{1cm}|}

\hline
\textbf{Search Engines} & 
\textbf{RBAC} & 
\textbf{role based access control} & 
\textbf{role-based access control} & 
\textbf{Total}
\\\hline

Google Scholar & 651 & 213 & 435 & 1299 \\\hline
ACM Portal & 500 & 20 & 720 & 1240 \\\hline
IEEExplore & 200 & 40 & 230 & 470 \\\hline
CiteSeerX & 100 & 100 & 150 & 350 \\\hline
 &  &  &  & \\\hline
Totals & 1451 & 373 & 1535 & 3359 \\\hline
Combined &  &  &  & \textbf{1716} \\\hline

\end{tabular}
\end{center}

@@

P1: Describe data extraction process, including questions each reviewer based note taking off of

P2: Describe how from the data extraction process cateogories of extension became evident

P3: Describe how those categories were deemed relevant and a way to organize going forward

P4: Describe the categories, show how each paper and their category

P5: Discuss how the categories are inter-related and show diagram of relationships

After selection of primary sources, the next step was to extract data from each paper that pertained to our research questions in order to look for trends.  The first step was to take the individual data generated from the final elimination round and organize this information around the research questions.  During the paper reading round, and resulting data, the fact that the papers were logically falling into a number of categorizations became evident.  Thus, the first step undertaken was to answer the question of what categories exist for the RBAC model extensions and what papers feel into what categories.  The result of this paper categorization is displayed below:

\begin{center}
\begin{tabular}{|p{12.5cm}|p{3cm}|}
\hline
\textbf{Paper} & \textbf{Category}
\\\hline
Alam, M. and Hafner, M. and Breu, R. 2006 \cite{alam06:constraint} & Constraint \\\hline
Tzelepi, Sofia K. and Koukopoulos, Dimitrios K. and Pangalos, George 2001 \cite{tzelepi01:flexible} & Context \\\hline
Haibo, SHEN and Fan, HONG 2005 \cite{haibo05:context} & Context \\\hline
Damian G. Cholewka and Reinhardt A. Botha and Jan H. P. Eloff 2000 \cite{cholewka00:acontext-sensitive} & Context \\\hline
Huang, X. and Wang, H. and Chen, Z. and Lin, J. 2006 \cite{huang06:pervasive} & Context \\\hline
Motta, G.H.M.B. and Furuie, S.S. 2003 \cite{motta03:contextual} & Context \\\hline
Bao, Y. and Song, J. and Wang, D. and Shen, D. and Yu, G. 2008 \cite{bao08:role} & Context \\\hline
Yamazaki, W. and Hiraishi, H. and Mizoguchi, F. 2004 \cite{yamazaki104:designing} & Context \\\hline
Jian-min, H. and Xi-yu, L. and Hui-qun, Y. and Jun, T. 2008 \cite{han08:extended} & Context \\\hline
Thein, N. and others 2011 \cite{thein2011leveraging} & Context \\\hline
Zou, D. and He, L. and Jin, H. and Chen, X. 2009 \cite{zou2009crbac} & Context \\\hline
Hasebe, K. and Mabuchi, M. and Matsushita, A. 2010 \cite{hasebe10:capability} & Delegation \\\hline
Zhang, Z. and Zhang, X. and Sandhu, R. 2006 \cite{zhang06:collaborative} & Organizational \\\hline
Ni, Q. and Trombetta, A. and Bertino, E. and Lobo, J. 2007 \cite{ni2010privacy} & Privacy \\\hline
Masoumzadeh, A. and Joshi, J. 2008 \cite{masoumzadeh2008purbac} & Privacy \\\hline
Zhao, Y. and Zhao, Y. and Lu, H. 2008 \cite{zhao2008flexible} & Resource \\\hline
Bertino, E. and Catania, B. and Damiani, M.L. and Perlasca, P. 2005 \cite{damian2007geo} & Spatial \\\hline
Ray, I. and Kumar, M. and Yu, L. 2006 \cite{ray07:spatio} & Spatial \\\hline
Hansen, F. and Oleshchuk, V. 2003 \cite{hansen2003spatial} & Spatial \\\hline
Aich, S. and Sural, S. and Majumdar, A. 2007 \cite{aich07:STARBAC} & Spatio-Temporal \\\hline
Chen, L. and Crampton, J. 2008 \cite{chen08:spatio-temporal} & Spatio-Temporal \\\hline
Samuel, A. and Ghafoor, A. and Bertino, E. 2007 \cite{samuel07:spatio-temporal} & Spatio-Temporal \\\hline
Chandran, S. and Joshi, J. 2005 \cite{chandran05:llt} & Spatio-Temporal \\\hline
Ray, I. and Toahchoodee, M. 2007 \cite{ray07:spatio} & Spatio-Temporal \\\hline
Aich, S. and Mondal, S. and Sural, S. and Majumdar, A. 2009 \cite{aich09:role} & Spatio-Temporal \\\hline
Yao, L. and Kong, X. and Xu, Z. 2008 \cite{yao2008task} & Task \\\hline
ZHANG, S. and CHEN, X. and HOU, G. 2009 \cite{zhou2007team} & Task \\\hline
Oh, S. and Park, S. 2003 \cite{oh2003task} & Task \\\hline
Joshi, J.B.D. and Bertino, E. and Latif, U. and Ghafoor, A. 2005 \cite{joshi05:generalized} & Temporal \\\hline
\end{tabular}
\end{center}

>>>>

@@

After selection of primary sources, the next step was to extract data from each paper that pertained to our research questions in order to look for trends.  The first step was to take the individual data generated from the final elimination round and organize this information around the research questions.  During the paper reading round, and resulting data, the fact that the papers were logically falling into a number of categorizations became evident.  Thus, the first step undertaken was to answer the question of what categories exist for the RBAC model extensions and what papers feel into what categories.  Each paper deemed a primary source and the papers associated categorization are shown in Table \ref{tab:categorization}.
